---
title: Modified Score Test for Linear Mixed Models
author: Karl Oskar Ekvall
output: html_document
---

# Installation

The package can be installed from GitHub using devtools.

```{r}
# devtools::install_github("koekvall/lmmstest")
library(lmmstest)
```


# Setting

Suppose a vector of responses $Y\in \mathbb{R}^n$, matrix of predictors $X\in
\mathbb{R}^{n\times p}$, and design matrix $Z \in \mathbb{R}^{n\times q}$
satisfy, for some $\beta \in \mathbb{R}^p$,
$$
  Y = X\beta + ZU + E,
$$
where $U = [U_1, \dots, U_q]^T$ is a multivariate normal vector of random
effects with mean zero and the error vector $E = [E_1, \dots, E_n]^T$ is
multivariate normal with mean zero and covariance matrix $\sigma^2 I_n$,
independent of $U$. Assume further that the elements of $U$ are independent and
that, for every $j \in \{1, \dots, q\}$, there is a $k \in \{1, \dots, d\}$ such
that $\mathrm{var}(U_j) = \lambda_k^2$. That is, the elements of $\lambda =
[\lambda_1, \dots, \lambda_d]^T \in [0, \infty)^d$ are the standard deviations
of the random effects. Several random effects can have the same variance, in
which case $d < q$.

# Background

Many popular tests perform poorly when one or more of the scale parameters
$\sigma, \lambda_1, \dots, \lambda_d$ are near or equal to zero. One way to
address this problem is to notice that the Fisher information is singular
when a scale parameter is zero. In particular, the element of the score
function corresponding to such a scale parameter is equal to zero almost
surely. The modified score test addresses this by replacing such elements,
which are first order partial derivatives of the log-likelihood, by second
order partial derivatives. When no scale parameter is equal to zero the
modified score test statistic is the same as the usual score test statistic
standardized by expected Fisher information; when one or more scale
parameters are equal to zero the modified statistic can be defined as a limit
of the usual one.  In the linear mixed model, this limit has an
analytical expression which facilitates our implementation.

# Example

Suppose that for observational unit $i = 1, \dots, N$ and time
$t = 1, \dots, N_T$,
$$
  Y_{i, t} = \beta_1 + \beta_2 X_{i, t} + U_{1, i} + U_{2, i} X_{i, t} + E_{i, t}.
$$
This model can be written in stacked form by letting $Y = [Y_{1, 1}, \dots,
Y_{N, N_T}]^T$ and defining $X$, $Z$, $U$, and $E$ accordingly. In particular,
$Z \in \mathbb{R}^{n \times 2N}$ with $n = NN_T$. The random effects $U_{1, i}
\sim \mathcal{N}(0, \lambda_1)$ and $U_{2, i} \sim \mathcal{N}(0, \lambda_2)$
are, respectively, a random intercept and a random slope, shared by observations
from the same unit. In stacked notation,
\[
  U = \begin{bmatrix} U_{1, 1} \\ U_{1, 2} \\ \vdots \\ U_{2, N}\end{bmatrix}
  \sim \mathcal{N}\left(0, \mathrm{diag}(\lambda_1, \lambda_2) \otimes
  I_N\right).
\]

Consider testing the null hypothesis $(\sigma, \lambda_1, \lambda_2) = (1, 0.5,
0)$. When $\lambda_2 = 0$, the Fisher information is singular and common test
statistics are unreliable. The following code generates one realization from the
model with true parameters consistent with the null hypothesis and tests the
null hypothesis using the proposed procedure. The predictors are drawn
independently from a uniform distribution on $[-1, 2]$.

```{r}
# For replicability
set.seed(2)

# Sample size
N <- 30
N_T <- 10
n <- N * N_T

# True parameters
sigma <- 1
lambda <- c(0.5, 0)
Beta <- c(1, 1)

# Generate matrix of data where rows index units and columns index time
XX <- matrix(runif(N * N_T, -1, 2) , nrow = N, ncol = N_T)
U1 <- rnorm(N, sd = lambda[1])
U2 <- rnorm(N, sd = lambda[2])
E <- matrix(rnorm(N * N_T, sd = sigma), nrow = N, ncol = N_T)
Y <-  Beta[1] + Beta[2] * XX + E
for(ii in 1:N){
  Y[ii, ] <- Y[ii, ] + U1[ii] + U2[ii] * XX[ii, ]
}

# Put data in tall format and add unit and time factors
ex_data <- data.frame(c(t(Y)), c(t(XX)))
ex_data$time <- rep(1:N_T, N)
ex_data$unit <- as.factor(rep(1:N, each = N_T))
names(ex_data)[1:2] <- c("y", "x")

# Inspect data
head(ex_data, 3)
tail(ex_data, 3)

# Create model matrices
X <- model.matrix(~x, data = ex_data)
Z <- model.matrix(~0 + unit + unit:x, data = ex_data)

# The first N columns of Z correspond to random intercepts and
# the next 30 to random slopes.
dim(Z)
lam_idx <- rep(1:2, each = N)
L <- diag(lambda[lam_idx], 2 * N)

# Get estimate of coefficient vector under the null hypothesis.
# Since Sigma is known under the null hypothesis, this estimate is the
# generalized least squares estimate.
Sigma_null <- diag(sigma^2, n) + Z %*% L^2 %*% t(Z)
Beta_null <- c(qr.solve(crossprod(X, qr.solve(Sigma_null, X)),
                      crossprod(X, qr.solve(Sigma_null, ex_data$y))))
Beta_null

# Compute modified score test statistic for elements 3:5 of theta =
# c(Beta, sigma, lambda[1], lambda_2), i.e. for (sigma, lambda[1], lambda_2).
# The supplied arguments for  (sigma, lambda[1], lambda_2) are the true values,
# so the null hypothesis is true.
lmmstest::score_test(y = ex_data$y,
                     X = X,
                     Z = Z,
                     Beta = Beta_null,
                     sigma = sigma,
                     lambda = lambda,
                     lam_idx = lam_idx,
                     test_idx = 3:5)
```

# Comparison to common test statistics

The following figure shows quantile-quantile plots for the modified score
(circles), likelihood ratio (triangles), and Wald (plus sign) test statistics
from a simulation. The theoretical quantiles are from a chi-square distribution
with 3 degrees of freedom. Data were generated according to the model from the
previous section assuming the null hypothesis $(\sigma, \lambda_1, \lambda_2) =
(1, 0, 0)$ (first plot), $(1, 0.01, 0.01)$ (second plot), or  $(1, 1, 1)$ (third
plot).

```{r, echo = F, fig.align='center', fig.width = 10, fig.height = 5}
sim_data <- readRDS("sims.Rds")
n_sims <- nrow(sim_data[[1]])
pp <- ppoints(n_sims)
par(mfrow = c(1, 3))
plot_dat <- sim_data[[17]]
plot(x = qchisq(pp, df = 3),
     y = quantile(plot_dat[, "chi_sq"], pp),
     xlab = "theoretical quantiles",
     ylab = "empirical quantiles")
points(x = qchisq(pp, df = 3),
       y = quantile(plot_dat[, "lrt_chi_sq"], pp),
       pch = 2)
points(x = qchisq(pp, df = 3),
       y = quantile(plot_dat[, "wald_chi_sq"], pp),
       pch = 3)
abline(a = 0, b = 1, col = "red")

plot_dat <- sim_data[[18]]
plot(x = qchisq(pp, df = 3),
     y = quantile(plot_dat[, "chi_sq"], pp),
     xlab = "theoretical quantiles",
     ylab = "empirical quantiles")
points(x = qchisq(pp, df = 3),
       y = quantile(plot_dat[, "lrt_chi_sq"], pp),
       pch = 2)
points(x = qchisq(pp, df = 3),
       y = quantile(plot_dat[, "wald_chi_sq"], pp),
       pch = 3)
abline(a = 0, b = 1, col = "red")

plot_dat <- sim_data[[24]]
plot(x = qchisq(pp, df = 3),
     y = quantile(plot_dat[, "chi_sq"], pp),
     xlab = "theoretical quantiles",
     ylab = "empirical quantiles")
points(x = qchisq(pp, df = 3),
       y = quantile(plot_dat[, "lrt_chi_sq"], pp),
       pch = 2)
points(x = qchisq(pp, df = 3),
       y = quantile(plot_dat[, "wald_chi_sq"], pp),
       pch = 3)
abline(a = 0, b = 1, col = "red")
```

When no scale parameter is zero, classical theory says all three test statistics
are asymptotically chi-square distributed with 3 degrees of freedom. The third
plot indicates this asymptotic distribution is a good approximation when scale
parameters are substantially different from zero.

The second plot shows the asymptotic distribution is a good approximation for
the proposed procedure but not for the other two when scale parameters are near
zero.

When some scale parameters are zero, the asymptotic distributions of the
likelihood ratio and Wald test statistics are no longer chi-square
distributions, and accordingly the first plot shows poor agreement between the
quantiles. By contrast, the modified score test has an asymptotic chi-square
distribution also in this setting, and the first plot shows the asymptotic
distribution provides a good approximation.


